---
title: "STA 602 Homework 6"
author: "Yutong Shao"
date: "2023-03-06"
output: 
     pdf_document
      # includes: 
      #     in_header: custom2.tex
font-size: 8px
editor_options: 
  markdown: 
    wrap: 72
---

# Problem Setting

Researchers are studying the length of life (lifetime) following a
particular medical intervention, such as a new surgical treatment for
heart disease, where the study consists of 12 patients. Specifically,
the number of years before death for each is
$$3.4, 2.9, 1.2+, 1.4, 3.2, 1.8, 4.6, 1.7+, 2.0+, 1.4+, 2.8, 0.6+$$
where the $+$ indicates that the patient was alive after $x$ years, but
the researchers lost contact with the patient after that point in time.

One way we can model this data is in the following way:

$$
\begin{aligned}
  &X_i = 
\begin{cases}
Z_i, & Z_i \leq c_i\\
c_i, & Z_i > c_i
\end{cases}\\
    &Z_1,\ldots,Z_n|\theta\,\stackrel{iid}{\sim} \,Gamma(r,\theta)\\
    &\theta\sim Gamma(a, b)
\end{aligned} 
$$

where $a$, $b$, and $r$ are known. In addition, we know:

```{=tex}
\begin{itemize}
\item $c_i$ is the censoring time for patient $i$, which is fixed, but known only if censoring occurs.
\item $X_i$ is the observation
\begin{itemize}
\item if the lifetime is less than $c_i$ then we get to observe it ($X_i = Z_i$), 
\item otherwise
        all we know is the lifetime is greater than $c_i$ ($X_i = c_i$).
        \end{itemize}
    \item $\theta$ is the parameter of interest---the rate parameter for the lifetime distribution.
    \item $Z_i$ is the lifetime for patient $i$, however, this is not directly observed.
    \end{itemize}
```
The probability density function (pdf) associated consists of two point
masses:one at $Z_i$ and one at $c_i$. The formula is

$$
\begin{aligned}
p(x_i|z_i) = \mathbb{1}(x_i = z_i)\mathbb{1}(z_i \leq c_i) + \mathbb{1}(x_i=c_i)\mathbb{1}(z_i > c_i).
\end{aligned}
$$

Now we can easily find the full conditionals (derived in class and
reproduced below). Notice that $z_i$ is conditionally independent of
$z_j$ given $\theta$ for $i \neq j$. This implies that $x_i$ is
conditionally independent of $x_j$ given $z_i$ for $i \neq j$. Now we
have

$$
\begin{aligned}
p(z_i|z_{-i},x_{1:n},\theta) &= p(z_i|x_i,\theta) \\
&\underset{z_i}{\propto} p(z_i,x_i,\theta) \\
&= p(\theta)p(z_i|\theta)p(x_i|z_i,\theta) \\
&\underset{z_i}{\propto} p(z_i|\theta)p(x_i|z_i,\theta) \\
&= p(z_i|\theta)p(x_i|z_i).
\end{aligned}
$$

There are now two cases to consider. If $x_i \neq c_i$, then
$p(z_i|\theta)p(x_i|z_i)$ is only non-zero when $z_i = x_i$. The density
devolves to a point mass at $x_i$. This corresponds to the case where
$z_i$ is observed, so $x_i$ is the observed value and we should always
sample this value. Practically speaking, we do not sample this value
when running the Gibbs sampler.

If $x_i = c_i$, then the density becomes
$p(x_i|z_i) = \mathbb{1}(z_i > c_i)$, so

$$
\begin{aligned}
p(z_i|\cdots) \propto p(z_i|\theta)\mathbb{1}(z_i>c_i),
\end{aligned}
$$

which is a truncated Gamma.

For the Gibbs sampler, we will use the current value of $\theta$ to
impute the censored data. We will sample from the truncated gamma using
a modified version of the iverse CDF trick. For the censored values of
$Z_i$ we know $c_i$. If we know $\theta$ (which we will in a Gibbs'
sampler), we know the distribution of $Z_i|\theta \sim Gamma(r,\theta)$.
Let $F$ be the CDF of this distribution. Suppose we truncate this
distribution to $(c,\infty)$. The new CDF is

$$
\begin{aligned}
P(Z_i < z) &= \frac{F(z) - F(c)}{1 - F(c)}.
\end{aligned}
$$

Therefore $Y$ is a sample from the truncated Gamma, as desired.

In the actual code for the Gibbs' sampler we do not sample the observed
values. We simply impute the censored values using the method above.

You will find code below (that is also taken from class ) that will help
you with the remainder of the problem.

1.  (5 points) Write code to produce trace plots and running average
    plots for the censored values for 200 iterations. Do these
    diagnostic plots suggest that you have run the sampler long enough?
    Explain.

2.  (5 points) Now run the chain for 10,000 iterations and update your
    diagnostic plots (traceplots and running average plots). Report your
    findings for both traceplots and the running average plots for
    $\theta$ and the censored values. Do these diagnostic plots suggest
    that you have run the sampler long enough? Explain.

3.  (5 points) Give plots of the estimated density of
    $\theta \mid \cdots$ and $z_9 \mid \cdots$. Be sure to give brief
    explanations of your results and findings. (Present plots for 10,000
    iterations).

4.  (5 points) Finally, let's suppose that $r=10,a=1,b=100.$ Do the
    posterior densities in part (c) change for $\theta \mid \cdots$ and
    $z_9 \mid \cdots?$ Do the associated posterior densities change when
    $r=10, a=100,b=1?$ Please provide plots and an explanation to back
    up your answer. (Use 10,000 iterations for the Gibbs sampler).

# Answers

## Question (a)

(a) (5 points) Write code to produce trace plots and running average
    plots for the censored values for 200 iterations. Do these
    diagnostic plots suggest that you have run the sampler long enough?
    Explain.

```{r}
knitr::opts_chunk$set(cache=FALSE)
library(xtable)

set.seed(2023)

# Samples from a truncated gamma with
# truncation (t, infty), shape a, and rate b
# Input: t,a,b
# Output: truncated Gamma(a,b)
sampleTrunGamma <- function(t, a, b){
  # This function samples from a truncated gamma with
  # truncation (t, infty), shape a, and rate b
  p0 <- pgamma(t, shape = a, rate = b)
  x <- runif(1, min = p0, max = 1)
  y <- qgamma(x, shape = a, rate = b)
  return(y)
}

# Gibbs sampler for censored data
# Inputs:
  # this function is a Gibbs sampler
  # z is the fully observe data
  # c is censored data
  # n.iter is number of iterations
  # init.theta and init.miss are initial values for sampler
  # r,a, and b are parameters
  # burnin is number of iterations to use as burnin
# Output: theta, z
sampleGibbs <- function(z, c, n.iter, init.theta, init.miss, r, a, b, burnin = 1){

  z.sum <- sum(z)
  m <- length(c)
  n <- length(z) + m
  miss.vals <- init.miss 
  res <- matrix(NA, nrow = n.iter, ncol = 1 + m)
  for (i in 1:n.iter){
    var.sum <- z.sum + sum(miss.vals)
    theta <- rgamma(1, shape = a + n*r, rate = b + var.sum)
    miss.vals <- sapply(c, function(x) {sampleTrunGamma(x, r, theta)})
    res[i,] <- c(theta, miss.vals)
  } 
  return(res[burnin:n.iter,])
} 

# set parameter values
r <- 10
a <- 1
b <- 1
# input data
z <- c(3.4,2.9,1.4,3.2,1.8,4.6,2.8)
c <- c(1.2,1.7,2.0,1.4,0.6)


n.iter <- 200
init.theta <- 1
init.missing <- rgamma(length(c), shape = r, rate = init.theta)
# run sampler
res <- sampleGibbs(z, c, n.iter, init.theta, init.missing, r, a, b)
```

In figure \ref{fig:trace-theta} and \ref{fig:trace-z} we see traceplots
for 200 iterations of the Gibbs sampler. It is difficult to tell whether
or not the sampler has failed to converge, thus, we turn to running
average plots.

```{r fig.width=7,fig.height=4,fig.cap="\\label{fig:trace-theta}Traceplot of theta"}
plot(1:n.iter, res[,1], pch = 16, cex = .35,
     xlab = "Iteration", ylab = expression(theta),
     main = expression(paste("Traceplot of ", theta)))
```

In figures \ref{fig:run-theta} and \ref{fig:run-z} we see running
average plots for 200 iterations of the Gibbs sampler, where from all of
these it is clear that after 200 iterations the sampler is having mixing
issues, and **should be run for longer** to check that "it has not
failed to converge."

```{r fig.width=7,fig.height=6,fig.cap="\\label{fig:trace-z}Traceplot of $z_3, z_8, z_9, z_{10}, z_{12}.$"}
missing.index <- c(3,8,9,10,12)
par(mfrow=c(3,2))
for (ind in missing.index){
  x.lab <- bquote(z[.(ind)])
  plot(1:n.iter, res[,which(missing.index == ind)], pch = 16, cex = .35,
       xlab = "Iteration", ylab = x.lab,
       main = bquote(paste("Traceplot of ", .(x.lab)))) 
}
plot.new()
```

```{r}
# get running averages
run.avg <- apply(res, 2, cumsum)/(1:n.iter)
```

```{r fig.width=7,fig.height=4,fig.cap="\\label{fig:run-theta}Running average plot of theta"}
plot(1:n.iter, run.avg[,1], type = "l",
     xlab = "Iteration", ylab = expression(theta),
     main = expression(paste("Running Average Plot of ", theta)))
```

```{r fig.width=7,fig.height=6,fig.cap="\\label{fig:run-z}Running average plots of $z_3, z_8, z_9, z_{10}, z_{12}.$"}
par(mfrow=c(3,2))
missing.index <- c(3,8,9,10,12)
for (ind in missing.index){
  x.lab <- bquote(z[.(ind)])
  plot(1:n.iter, run.avg[,which(missing.index == ind)], type = "l",
       xlab = "Iteration", ylab = x.lab,
       main = bquote(paste("Running Average Plot of ", .(x.lab)))) 
}
plot.new()
```

Figures \ref{fig:density-theta} and \ref{fig:density-z} do not provide
meaningful inference at this point since the sampler **has not been run
long enough**.

```{r}
# calculate confidence interval
quantile(res[,1], c(0.025, 0.975))
```

```{r fig.width=7,fig.height=4,fig.cap="\\label{fig:density-theta}Estimated posterior density of theta"}
# density plots

plot(density(res[,1]), xlab = expression(theta), 
     main = expression(paste("Density of ", theta)))
abline(v = mean(res[,1]), col = "red")
abline(v = c(2.60, 4.09), col = "blue")
```

```{r}
quantile(res[,4], c(0.025, 0.975))
```

```{r fig.width=7,fig.height=4,fig.cap="\\label{fig:density-z}Estimated posterior density of $z_9$ (posterior mean in red)."}
plot(density(res[,4]), xlab = expression(z[9]),
     main = expression(paste("Density of ", z[9])))
abline(v = mean(res[,4]), col = "red")
abline(v = c(2.08, 5.65), col = "blue")
```

## Question (b)

(b) (5 points) Now run the chain for 10,000 iterations and update your
    diagnostic plots (traceplots and running average plots). Report your
    findings for both traceplots and the running average plots for
    $\theta$ and the censored values. Do these diagnostic plots suggest
    that you have run the sampler long enough? Explain.

```{r}
n_iter_2 <- 10000
init.theta <- 1
init.missing <- rgamma(length(c), shape = r, rate = init.theta)
# run sampler
res_2 <- sampleGibbs(z, c, n_iter_2, init.theta, init.missing, r, a, b)


```

Traceplot of $\theta$ and censored values

```{r fig.width=7,fig.height=4, fig.cap="\\label{fig:trace-theta-10000}Traceplot of theta"}
plot(1:n_iter_2, res_2[,1], pch = 16, cex = .35,
     xlab = "Iteration", ylab = expression(theta),
     main = expression(paste("Traceplot of ", theta, " with 10000 iterations")))
```

```{r fig.width=7,fig.height=6,fig.cap="\\label{fig:trace-z-10000}Traceplot of $z_3, z_8, z_9, z_{10}, z_{12}.$"}
missing.index <- c(3,8,9,10,12)
par(mfrow=c(3,2))
for (ind in missing.index){
  x.lab <- bquote(z[.(ind)])
  plot(1:n_iter_2, res_2[,which(missing.index == ind)], pch = 16, cex = .35,
       xlab = "Iteration", ylab = x.lab,
       main = bquote(paste("Traceplot of ", .(x.lab), " with 10000 iterations"))) 
}
plot.new()
```

Running Average plots of $\theta$ and censored values.

```{r}
# get running averages
run_avg_2 <- apply(res_2, 2, cumsum)/(1:n_iter_2)
```

```{r fig.width=7,fig.height=4,fig.cap="\\label{fig:run-theta-10000}Running average plot of theta"}
plot(1:n_iter_2, run_avg_2[,1], type = "l",
     xlab = "Iteration", ylab = expression(theta),
     main = expression(paste("Running Average Plot of ", theta, " with 10000 iterations")))
```

```{r fig.width=7,fig.height=6,fig.cap="\\label{fig:run-z-10000}Running average plots of $z_3, z_8, z_9, z_{10}, z_{12}.$"}
par(mfrow=c(3,2))
missing.index <- c(3,8,9,10,12)
for (ind in missing.index){
  x.lab <- bquote(z[.(ind)])
  plot(1:n_iter_2, run_avg_2[,which(missing.index == ind)], type = "l",
       xlab = "Iteration", ylab = x.lab,
       main = bquote(paste("Running Average Plot of ", .(x.lab), " with 10000 iterations"))) 
}
plot.new()
```

From the Traceplots \ref{fig:trace-theta-10000} and
\ref{fig:trace-z-10000}, it is not easy to infer whether they have
converged or not. But the running average plots
\ref{fig:run-theta-10000} and \ref{fig:run-z-10000} clearly show that
both $\theta$ and censored values have converged after 2000 iterations,
because the lines tend to be flat as we keep iterating. Thus, we already
run long enough for convergence.

## Question (c)

(c) (5 points) Give plots of the estimated density of
    $\theta \mid \cdots$ and $z_9 \mid \cdots$. Be sure to give brief
    explanations of your results and findings. (Present plots for 10,000
    iterations).

```{r fig.width=7,fig.height=4,fig.cap="\\label{fig:density-theta-10000}Estimated posterior density of theta"}
# density plots
quantile(res_2[,1], c(0.025, 0.975))
plot(density(res_2[,1]), xlab = expression(theta), 
     main = expression(paste("Density of ", theta, " with 10000 iterations")))
abline(v = mean(res_2[,1]), col = "red")
abline(v = c(2.57, 4.05), col = "blue")
```

```{r fig.width=7,fig.height=4,fig.cap="\\label{fig:density-z-10000}Estimated posterior density of $z_9$ (posterior mean in red)."}
quantile(res_2[,4], c(0.025, 0.975))
plot(density(res_2[,4]), xlab = expression(z[9]),
     main = expression(paste("Density of ", z[9])))
abline(v = mean(res_2[,4]), col = "red")
abline(v = c(2.07, 5.53), col = "blue")
```

The first graph shows the estimated density of $\theta$ is an updated
Gamma distribution. This represents the approximated posterior density
of an individual's lifetime before death. The posterior mean shifted
slightly to the left as compared to the 200-iteration case.

The second graph is the estimated density of $z_9$, which is the imputed
value of lifetime of individual 9. Here the density of $z_9$ is more
skewed as the mode shifted more to left than the mean, comparing with
the 200-iteration case. Thus, increasing the number of iterations is
necessary to achieve reliable distributions.

## Question (d)

(d) (5 points) Finally, let's suppose that $r=10,a=1,b=100.$ Do the
    posterior densities in part (c) change for $\theta \mid \cdots$ and
    $z_9 \mid \cdots?$ Do the associated posterior densities change when
    $r=10, a=100,b=1?$ Please provide plots and an explanation to back
    up your answer. (Use 10,000 iterations for the Gibbs sampler).

### $r=10, a=1, b=100$

```{r}
# new parameter values
r <- 10
a <- 1
b1 <- 100

# run sampler
res_3 <- sampleGibbs(z, c, n_iter_2, init.theta, init.missing, r, a, b1)
quantile(res_3[,1], c(0.025, 0.975))
quantile(res_3[,4], c(0.025, 0.975))
```

```{r fig.width=7,fig.height=4,fig.cap="\\label{fig:density-theta-10000}Estimated posterior density of theta"}
# density plots

plot(density(res_3[,1]), xlab = expression(theta), 
     main = expression(paste("Density of ", theta, " with 10000 iterations, b=100")))
abline(v = mean(res_3[,1]), col = "red")
abline(v = c(0.46, 0.73), col = "blue")
```

```{r fig.width=7,fig.height=4,fig.cap="\\label{fig:density-z-10000}Estimated posterior density of $z_9$ (posterior mean in red)."}
plot(density(res_3[,4]), xlab = expression(z[9]),
     main = expression(paste("Density of ", z[9], ", 10000 iterations, b=100")))
abline(v = mean(res_3[,4]), col = "red")
abline(v = c(8.06, 30.91), col = "blue")
```

### $r=10, a=100, b=1$

```{r}
# new parameter values
r <- 10
a1 <- 100
b <- 1

# run sampler
res_4 <- sampleGibbs(z, c, n_iter_2, init.theta, init.missing, r, a1, b)
quantile(res_4[,1], c(0.025, 0.975))
quantile(res_4[,4], c(0.025, 0.975))
```

```{r fig.width=7,fig.height=4,fig.cap="\\label{fig:density-theta-10000}Estimated posterior density of theta"}
# density plots
plot(density(res_4[,1]), xlab = expression(theta), 
     main = expression(paste("Density of ", theta, ", 10000 iterations, a=100")))
abline(v = mean(res_4[,1]), col = "red")
abline(v = c(6.31, 8.34), col = "blue")
```

```{r fig.width=7,fig.height=4,fig.cap="\\label{fig:density-z-10000}Estimated posterior density of $z_9$ (posterior mean in red)."}
plot(density(res_4[,4]), xlab = expression(z[9]),
     main = expression(paste("Density of ", z[9], ", 10000 iterations, a=100")))
abline(v = mean(res_4[,4]), col = "red")
abline(v = c(2.01, 3.00), col = "blue")
```

The posterior density changes with different parameters, and the 95%
confidence interval is very different.

-   When $b=100$ and others remains the same, the posterior values of
    $\theta$ decreases with most values concentrate on 0.6. While the
    posterior of $z_9$ increases a lot, with point mass around 15, which
    is not quite reasonable comparing other individual's lifetime.

-   When $a=100$ and others remains the same, the posterior values of
    $\theta$ increases with most values concentrate on around 7. While
    the posterior of $z_9$ slightly decreases, with point mass on around
    2.1. And the density around $z_9\in (0,2)$ is very close to zero,
    meaning the lifetime of the 9th patient with censored data will
    probably have 2.2 to 2.5 years of living after treatment.

Thus, different choices of parameters will greatly influence the
posterior density estimation. We should be careful on choosing the prior
parameters.

# References

Code was adapted from course materials provided by Prof. Rebecca C.
Steorts.
